---
title: "Data Processing and Visualization in R"
date: "`r Sys.Date()`"
output: html_document
---

# Install tidyverse
``` {r install_packages, echo=FALSE}
# Install necessary packages
#if (!requireNamespace("tidyverse", quietly = TRUE)) {
#    install.packages("tidyverse")
#}
```

# Load tidyverse

``` {r install_packages2, echo=TRUE}
library(tidyverse)
```


# Load and prepare the data
``` {r load_data, echo=TRUE}
data <- read_csv("01_data/PXD040621_peptides.csv", show_col_types = FALSE)
head(data)
```

## Data transformation

log2 transformations are common for lognormal distributed data
``` {r transform_data, echo=TRUE}
data <- data %>%
    mutate(Intensity = log2(Intensity))
```

## Data aggregation

Aggregate the peptide intensities to protein intensities.
We use the median of the peptide intensities for each protein.
Also, let's shorten sample names for better readability.

``` {r aggregate_data, echo=TRUE}
protein_data <- data %>%
    group_by(ProteinName, Reference) %>%
    summarize(Intensity = median(Intensity, na.rm = TRUE), .groups = "drop") %>%
    mutate(Reference = sapply(strsplit(Reference, "_"), function(x) paste(x[4:6], collapse = "_"))) %>%
    mutate(Reference = str_remove(Reference, "^Ecoli_"))
```

## Remove contaminants

Remove the contaminant proteins which were added to the fasta file used in the data processing.
Contaminant proteins are e.g. creation which gets into the sample from the human skin/hair
when sample is prepared.

``` {r remove_contaminants, echo=TRUE}
protein_data <- protein_data %>%
    filter(!str_detect(ProteinName, "CON_"))

# Check the data after removing contaminants.
head(protein_data, n = 20)
```

## Cleaning names
Split the ProteinName column into 'Identifier', 'Source', 'ProteinName' and 'GeneName'
After splitting, we also remove the '_ECOLI' suffix from the GeneName column.


``` {r parse_data, echo=TRUE}
protein_data_parsed <- protein_data %>%
    separate(ProteinName, into = c("Source", "Protein", "Gene"), sep = "\\|", extra = "drop", remove = FALSE) %>%
    mutate(Gene = str_remove(Gene, "_ECOLI")) %>%
    rename("Identifier" = "ProteinName")

# Finally, add a column with the experimental condition labels
protein_data_parsed <- protein_data_parsed %>%
    mutate(Label = if_else(str_detect(Reference, "DMSO"), "DMSO", "Sulforaphane"))
```

## Plot overall intensity distribution per replicate
```{r plot_distribution, echo=TRUE, fig.width=10, fig.height=6}
# Define colors for the labels
label_colors <- c("DMSO" = "#6a51a3", "Sulforaphane" = "#2ca25f")

# Start the plotting pipeline
protein_data_parsed %>%
    # Format Reference names for better readability in the plots
    mutate(Reference = case_when(
        Reference == "DMSO_rep1" ~ "DMSO 1",
        Reference == "DMSO_rep2" ~ "DMSO 2",
        Reference == "DMSO_rep3" ~ "DMSO 3",
        Reference == "DMSO_rep4" ~ "DMSO 4",
        Reference == "Suf_rep1" ~ "Sulforaphane 1",
        Reference == "Suf_rep2" ~ "Sulforaphane 2",
        Reference == "Suf_rep3" ~ "Sulforaphane 3",
        Reference == "Suf_rep4" ~ "Sulforaphane 4",
        TRUE ~ Reference
    )) %>%
    # Start ggplot, map global aesthetics (x axis and fill color)
    ggplot(., aes(x = Intensity, fill = Label)) +
    # Add histogram layer with 40 bins and white borders
    geom_histogram(bins = 40, color = "white") +
    # Add labels and titles
    labs(
        title = "Log2-transformed protein intensity distributions",
        x = "Log2 [Intensity]",
        y = "Count",
        fill = "Group"
    ) +
    # Adjust y-axis to make bars touch the x-axis
    scale_y_continuous(expand = c(0.01, 0)) +
    # Use a minimal plot theme to start the customization
    theme_minimal() +
    # Use the above defined colors for the fill aesthetic
    scale_fill_manual(values = label_colors) +
    # Create a separate facet (subplot) for each Reference (replicate)
    facet_wrap(~Reference, scales = "fixed", nrow = 2) +
    # Further customize the theme
    theme(
        axis.text = element_text(size = 12, color = "#666666"),
        axis.title = element_text(size = 13),
        plot.title = element_text(size = 14, face = "bold"),
        legend.title = element_blank(),
        legend.text = element_text(size = 12),
        legend.position = "bottom",
        strip.text = element_text(size = 12),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.border = element_rect(color = "#666666", fill = NA),
        axis.ticks = element_line(color = "#666666")
    )
```

### Perform Principal Component Analysis (PCA)
```{r pca, echo=TRUE, fig.width=10, fig.height=8}
# Reshape the data for PCA
pca_data <- protein_data_parsed %>%
    select(Identifier, Reference, Intensity) %>%
    pivot_wider(names_from = Identifier, values_from = Intensity) %>%
    column_to_rownames(var = "Reference") %>%
    select(where(~ !any(is.na(.)))) %>%
    select(where(~ var(.) > 0))

# Perform PCA
pca_result <- prcomp(pca_data, center = TRUE, scale. = TRUE)

# Get the percentage of variance explained by each principal component
pca_variance <- (pca_result$sdev^2) / sum(pca_result$sdev^2) * 100

# Create a data frame for plotting
pca_df <- as.data.frame(pca_result$x) %>%
    rownames_to_column(var = "Reference") %>%
    mutate(Label = if_else(str_detect(Reference, "DMSO"), "DMSO", "Sulforaphane"))

# Calculate convex hulls for each group (optional)
# These are used to draw polygons around the groups in the PCA plot
hulls <- pca_df %>%
    group_by(Label) %>%
    slice(chull(PC1, PC2)) %>%
    ungroup()

# Plot PCA results
pca_df %>%
    # Format Reference names for better readability in the plots
    mutate(Reference = case_when(
        Reference == "DMSO_rep1" ~ "DMSO 1",
        Reference == "DMSO_rep2" ~ "DMSO 2",
        Reference == "DMSO_rep3" ~ "DMSO 3",
        Reference == "DMSO_rep4" ~ "DMSO 4",
        Reference == "Suf_rep1" ~ "Sulforaphane 1",
        Reference == "Suf_rep2" ~ "Sulforaphane 2",
        Reference == "Suf_rep3" ~ "Sulforaphane 3",
        Reference == "Suf_rep4" ~ "Sulforaphane 4",
        TRUE ~ Reference
    )) %>%
    # Start ggplot, map global aesthetics (x and y axes, color, and point labels)
    ggplot(aes(x = PC1, y = PC2, color = Label, label = Reference)) +
    geom_point(size = 6) +
    # Draw convex hulls around the groups
    geom_polygon(
        data = hulls, aes(x = PC1, y = PC2, fill = Label, color = Label),
        alpha = 0.15, color = NA, show.legend = FALSE
    ) +
    # Add text labels to the points
    geom_text(vjust = -1.3, hjust = 0.5, size = 3.5, show.legend = FALSE) +
    # Add labels and titles
    labs(
        title = "Principal component analysis (PCA) of Log2-transformed protein intensities",
        x = paste0("PC1 (", round(pca_variance[1], 1), "% variance)"),
        y = paste0("PC2 (", round(pca_variance[2], 1), "% variance)"),
        color = "Group"
    ) +
    # Use the above defined colors for the color and fill aesthetics
    scale_color_manual(values = label_colors) +
    scale_fill_manual(values = label_colors) +
    # Use a minimal plot theme to start the customization
    theme_minimal() +
    # Further customize the theme
    theme(
        axis.text = element_text(size = 12, color = "#666666"),
        axis.title = element_text(size = 12.5),
        plot.title = element_text(size = 14, face = "bold"),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.position = "right",
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#666666", fill = NA),
        axis.ticks = element_line(color = "#666666")
    )
```

### Identify and plot proteins with highest fold-change between conditions
```{r fold_change, echo=TRUE, fig.width=12.5, fig.height=12}
# Compute fold-change per gene between Sulforaphane and DMSO
fold_changes <- protein_data_parsed %>%
    group_by(Gene, Label) %>%
    filter(n() >= 2) %>% # keep only labels with >=2 replicates
    ungroup() %>%
    group_by(Gene) %>%
    filter(n_distinct(Label) == 2) %>% # keep only proteins present in both groups
    summarize(
        DMSO = mean(Intensity[Label == "DMSO"], na.rm = TRUE),
        Sulforaphane = mean(Intensity[Label == "Sulforaphane"], na.rm = TRUE),
        .groups = "drop"
    ) %>%
    mutate(
        Log2FC = Sulforaphane - DMSO, # already log2-transformed
    ) %>%
    # Prepare a column indicating in which condition intensity is higher
    # This column will be used for faceting the plot later
    mutate(higher_in = case_when(
        Log2FC > 0 ~ "Higher in Sulforaphane",
        Log2FC < 0 ~ "Higher in DMSO",
        TRUE ~ "No change"
    )) %>%
    arrange(desc(Log2FC))

# Extract top 20 proteins with highest absolute Log2FC in each condition
top_proteins_DMSO <- fold_changes %>%
    filter(higher_in == "Higher in DMSO") %>%
    slice_max(order_by = abs(Log2FC), n = 20)

top_proteins_sulforaphane <- fold_changes %>%
    filter(higher_in == "Higher in Sulforaphane") %>%
    slice_max(order_by = abs(Log2FC), n = 20)

# Combine the top proteins from both conditions
top_proteins_both <- bind_rows(top_proteins_DMSO, top_proteins_sulforaphane)

# Prepare data for plotting
plot_data <- protein_data_parsed %>%
    group_by(Gene, Label) %>%
    summarize(
        Mean = mean(Intensity, na.rm = TRUE),
        SD = sd(Intensity, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    # Inner join the current data frame with the top proteins to keep only those
    inner_join(top_proteins_both[, c("Gene", "higher_in")], by = "Gene")

plot_data %>%
    # Start ggplot, map global aesthetics (x and y axes, color, and group)
    # Also, reorder genes based on their mean intensity for better visualization
    ggplot(aes(x = reorder(Gene, Mean), y = Mean, color = Label, group = Gene)) +
    # Add lines connecting the points of the same gene
    # (needs 'group' aesthetic for the correct behavior)
    geom_line(color = "#999999", linewidth = 1, show.legend = FALSE) +
    # Add "shadow" points representing the standard deviation.
    # They should be behind the main points, so add them first.
    geom_point(aes(size = SD),
        position = position_dodge(width = 0.5), alpha = 0.35,
        show.legend = TRUE
    ) +
    # Add points for each condition.
    geom_point(position = position_dodge(width = 0.5), size = 4, show.legend = FALSE) +
    # Refine size scale for the SD points for better visibility
    scale_size_continuous(range = c(6, 13), breaks = c(1, 2, 3)) +
    # Flip coordinates. What is mapped to x axis goes to y axis and vice versa.
    coord_flip() +
    # Add labels and titles
    labs(
        title = "Top 20 proteins with highest Log2FC between Sulforaphane and DMSO",
        x = "Gene symbol",
        y = "Mean Log2 [Intensity]",
        size = "Standard deviation",
        color = "Group"
    ) +
    # Use the above defined colors for the color aesthetic
    scale_color_manual(values = label_colors) +
    # Use a minimal plot theme to start the customization
    theme_minimal() +
    # Create a separate facet (subplot) for each condition where intensity is higher
    facet_wrap(~higher_in, scales = "free_y", nrow = 1) +
    # Further customize the theme
    theme(
        axis.text = element_text(size = 13, color = "#666666"),
        axis.title = element_text(size = 14),
        plot.title = element_text(size = 17, face = "bold"),
        legend.title = element_text(size = 13),
        legend.text = element_text(size = 13),
        legend.position = "right",
        legend.key.size = unit(1.2, "lines"),
        strip.text = element_text(size = 14),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#666666", fill = NA),
        axis.ticks = element_line(color = "#666666")
    ) +
    guides(color = guide_legend(
        override.aes = list(size = 5, alpha = 1)
    ))
```