# For Intermediate

Inspecting and summarizing the proteomics data.


```{r}
library(readr)
library(plotly)
library(tidyverse)
```

Inspecting and summarizing the proteomics data.

<div class="remember">
  <strong>📌 Remember:</strong> Load the library before starting the analysis.
</div>

**Load and prepare the data**
 
`read_csv` = reads a csv file

``` {r, echo=TRUE}
data <- read_csv("../data-01/PXD040621_peptides.csv", show_col_types = FALSE)
```

<div class="remember">
  <strong>📌 Remember:</strong> Remember: You can use the `DT` package to visualize the data.
</div>

```{r}
DT::datatable(
  data = head(data, 1000),  # show only the first 1000 rows
  rownames = FALSE,
  extensions = c('Buttons', 'Scroller'),
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv'),
    deferRender = TRUE,
    scrollX = TRUE,
    scrollY = 200,
    scroller = TRUE
  ),
  caption = 'proteomics metadata'
)
```

### Data transformation

`log2` transformations are common for lognormal distributed data

``` {r transform_data_, echo=TRUE}
data <- data %>%
    mutate(Intensity = log2(Intensity))
```

<div class="question">
  <strong>❓ Question:</strong> Do you remember what is the %>% doing in the code?
</div>


### Data aggregation

Aggregate the peptide intensities to protein intensities.
We use the median of the peptide intensities for each protein.
Also, let's shorten sample names for better readability.

``` {r aggregate_data_, echo=TRUE}
protein_data <- data %>%
    group_by(ProteinName, Reference) %>%
    summarize(Intensity = median(Intensity, na.rm = TRUE), .groups = "drop") %>%
    mutate(Reference = sapply(strsplit(Reference, "_"), function(x) paste(x[4:6], collapse = "_"))) %>%
    mutate(Reference = str_remove(Reference, "^Ecoli_"))
```


<div class="tip">
<strong>💡 Tip:</strong> Try to run line by line in the above code.
</div>


### Remove contaminants

Remove the contaminant proteins which were added to the fasta file used in the data processing.
Contaminant proteins are e.g. creation which gets into the sample from the human skin/hair
when sample is prepared.

``` {r remove_contaminants_, echo=TRUE}
protein_data <- protein_data %>%
    filter(!str_detect(ProteinName, "CON_"))

```

<div class="remember">
  <strong>📌 Remember:</strong> It is always a good idea to check the data after removing contaminants.
</div>

```{r}
head(protein_data, n = 10)
```


<div class="question">
  <strong>❓ Question:</strong> Can you think about a different way of inspecting the data?</div>


### Cleaning names

Split the ProteinName column into 'Identifier', 'Source', 'ProteinName' and 'GeneName'
After splitting, we also remove the '_ECOLI' suffix from the GeneName column.


``` {r parse_data_, echo=TRUE}
protein_data_parsed <- protein_data %>%
    separate(ProteinName, into = c("Source", "Protein", "Gene"), sep = "\\|", extra = "drop", remove = FALSE) %>%
    mutate(Gene = str_remove(Gene, "_ECOLI")) %>%
    rename("Identifier" = "ProteinName")
```


Finally, add a column with the experimental condition labels

```{r}
protein_data_parsed <- protein_data_parsed %>%
    mutate(Label = if_else(str_detect(Reference, "DMSO"), "DMSO", "Sulforaphane"))
```





```{r}
protein_data_parsed_mut <- protein_data_parsed %>%
    mutate(Reference = case_when(
        Reference == "DMSO_rep1" ~ "DMSO 1",
        Reference == "DMSO_rep2" ~ "DMSO 2",
        Reference == "DMSO_rep3" ~ "DMSO 3",
        Reference == "DMSO_rep4" ~ "DMSO 4",
        Reference == "Suf_rep1" ~ "Sulforaphane 1",
        Reference == "Suf_rep2" ~ "Sulforaphane 2",
        Reference == "Suf_rep3" ~ "Sulforaphane 3",
        Reference == "Suf_rep4" ~ "Sulforaphane 4",
        TRUE ~ Reference
    )) 
```



```{r}

plot_ly(
  data = protein_data_parsed_mut,
  x = ~Intensity,
  color = ~Reference,        # equivalent to fill in ggplot2
  type = "histogram"
) %>%
  layout(
    barmode = "stack", # "overlay" or "stack"
    title = "Protein Intensity Distribution by Label",
    xaxis = list(title = "Intensity"),
    yaxis = list(title = "Count")
  )

```


