# For Advanced


Inspecting and summarizing the proteomics data.

<div class="remember">
  <strong>📌 Remember:</strong> Load the library before starting the analysis.
</div>


``` {r install_packages2, echo=TRUE}
library(tidyverse)
```


**Load and prepare the data**
 
`read_csv` = reads a csv file

``` {r load_data, echo=TRUE}
data <- read_csv("../data-01/PXD040621_peptides.csv", show_col_types = FALSE)
```

<div class="remember">
  <strong>📌 Remember:</strong> Remember: You can use the `DT` package to visualize the data.
</div>

```{r}
DT::datatable(
  data = head(data, 1000),  # show only the first 1000 rows
  rownames = FALSE,
  extensions = c('Buttons', 'Scroller'),
  options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv'),
    deferRender = TRUE,
    scrollX = TRUE,
    scrollY = 200,
    scroller = TRUE
  ),
  caption = 'proteomics metadata'
)
```

### Data transformation

`log2` transformations are common for lognormal distributed data

``` {r transform_data, echo=TRUE}
data <- data %>%
    mutate(Intensity = log2(Intensity))
```

<div class="question">
  <strong>❓ Question:</strong> Do you remember what is the %>% doing in the code?
</div>


### Data aggregation

Aggregate the peptide intensities to protein intensities.
We use the median of the peptide intensities for each protein.
Also, let's shorten sample names for better readability.

``` {r aggregate_data, echo=TRUE}
protein_data <- data %>%
    group_by(ProteinName, Reference) %>%
    summarize(Intensity = median(Intensity, na.rm = TRUE), .groups = "drop") %>%
    mutate(Reference = sapply(strsplit(Reference, "_"), function(x) paste(x[4:6], collapse = "_"))) %>%
    mutate(Reference = str_remove(Reference, "^Ecoli_"))
```


<div class="tip">
<strong>💡 Tip:</strong> Try to run line by line in the above code.
</div>


### Remove contaminants

Remove the contaminant proteins which were added to the fasta file used in the data processing.
Contaminant proteins are e.g. creation which gets into the sample from the human skin/hair
when sample is prepared.

``` {r remove_contaminants, echo=TRUE}
protein_data <- protein_data %>%
    filter(!str_detect(ProteinName, "CON_"))

```

<div class="remember">
  <strong>📌 Remember:</strong> It is always a good idea to check the data after removing contaminants.
</div>

```{r}
head(protein_data, n = 10)
```


<div class="question">
  <strong>❓ Question:</strong> Can you think about a different way of inspecting the data?</div>


### Cleaning names

Split the ProteinName column into 'Identifier', 'Source', 'ProteinName' and 'GeneName'
After splitting, we also remove the '_ECOLI' suffix from the GeneName column.


``` {r parse_data, echo=TRUE}
protein_data_parsed <- protein_data %>%
    separate(ProteinName, into = c("Source", "Protein", "Gene"), sep = "\\|", extra = "drop", remove = FALSE) %>%
    mutate(Gene = str_remove(Gene, "_ECOLI")) %>%
    rename("Identifier" = "ProteinName")
```


Finally, add a column with the experimental condition labels

```{r}
protein_data_parsed <- protein_data_parsed %>%
    mutate(Label = if_else(str_detect(Reference, "DMSO"), "DMSO", "Sulforaphane"))
```


### Plot overall intensity distribution per replicate

Define colors for the labels

```{r plot_distribution, echo=TRUE, fig.width=10, fig.height=6}
label_colors <- c("DMSO" = "#6a51a3", "Sulforaphane" = "#2ca25f")

```

Start the plotting pipeline

  - Format Reference names for better readability in the plots

     Start ggplot, map global aesthetics (x axis and fill color)
     
Add histogram layer with 40 bins and white borders
    
Add labels and titles




```{r}
protein_data_parsed_mut <- protein_data_parsed %>%
    mutate(Reference = case_when(
        Reference == "DMSO_rep1" ~ "DMSO 1",
        Reference == "DMSO_rep2" ~ "DMSO 2",
        Reference == "DMSO_rep3" ~ "DMSO 3",
        Reference == "DMSO_rep4" ~ "DMSO 4",
        Reference == "Suf_rep1" ~ "Sulforaphane 1",
        Reference == "Suf_rep2" ~ "Sulforaphane 2",
        Reference == "Suf_rep3" ~ "Sulforaphane 3",
        Reference == "Suf_rep4" ~ "Sulforaphane 4",
        TRUE ~ Reference
    )) 
```



```{r}
library(plotly)

plot_ly(
  data = protein_data_parsed_mut,
  x = ~Intensity,
  color = ~Label,        # equivalent to fill in ggplot2
  type = "histogram",
  nbinsx = 30            # optional: number of bins
) %>%
  layout(
    barmode = "overlay", # "overlay" or "stack"
    title = "Protein Intensity Distribution by Label",
    xaxis = list(title = "Intensity"),
    yaxis = list(title = "Count")
  )

```


```{r}
ggplot(protein_data_parsed_mut, aes(x = Intensity, fill = Label)) +
  geom_histogram(bins = 40, color = "white") +
  labs(
    title = "Log2-transformed protein intensity distributions",
    x = "Log2 [Intensity]",
    y = "Count",
    fill = "Group"
  ) +
  scale_y_continuous(expand = c(0.01, 0)) +
  theme_minimal() +
  scale_fill_manual(values = label_colors) +
  facet_wrap(~Reference, scales = "fixed", nrow = 2)
```


## Perform Principal Component Analysis (PCA)


Reshape the data for PCA

```{r pca, echo=TRUE, fig.width=10, fig.height=8}
pca_data <- protein_data_parsed %>%
    select(Identifier, Reference, Intensity) %>%
    pivot_wider(names_from = Identifier, values_from = Intensity) %>%
    column_to_rownames(var = "Reference") %>%
    select(where(~ !any(is.na(.)))) %>%
    select(where(~ var(.) > 0))
```


** Perform the PCA**



```{r}
pca_result <- prcomp(pca_data, center = TRUE, scale. = TRUE)
```

**Get the percentage of variance explained by each principal component**


```{r}
pca_variance <- (pca_result$sdev^2) / sum(pca_result$sdev^2) * 100
```


**Create a data frame for plotting**


```{r}
pca_df <- as.data.frame(pca_result$x) %>%
    rownames_to_column(var = "Reference") %>%
    mutate(Label = if_else(str_detect(Reference, "DMSO"), "DMSO", "Sulforaphane"))

```


Calculate convex hulls for each group (optional)
These are used to draw polygons around the groups in the PCA plot

```{r}
hulls <- pca_df %>%
    group_by(Label) %>%
    slice(chull(PC1, PC2)) %>%
    ungroup()
```


Plot PCA results

```{r}
# Reuse of a code is always an smart idea
#protein_data_parsed_mut

    ggplot(pca_df, aes(x = PC1, y = PC2, color = Label, label = Reference)) +
    geom_point(size = 6) +
    # Draw convex hulls around the groups
    geom_polygon(
        data = hulls, aes(x = PC1, y = PC2, fill = Label, color = Label),
        alpha = 0.15, color = NA, show.legend = FALSE
    ) +
    geom_text(vjust = -1.3, hjust = 0.5, size = 3.5, show.legend = FALSE) +
    labs(
        title = "Principal component analysis (PCA) of Log2-transformed protein intensities",
        x = paste0("PC1 (", round(pca_variance[1], 1), "% variance)"),
        y = paste0("PC2 (", round(pca_variance[2], 1), "% variance)"),
        color = "Group"
    ) +
    scale_color_manual(values = label_colors) +
    scale_fill_manual(values = label_colors) +
    theme_minimal() +
    # Further customize the theme
    theme(
        axis.text = element_text(size = 12, color = "#666666"),
        axis.title = element_text(size = 12.5),
        plot.title = element_text(size = 14, face = "bold"),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.position = "right",
        panel.grid.minor = element_blank(),
        panel.border = element_rect(color = "#666666", fill = NA),
        axis.ticks = element_line(color = "#666666")
    )
```


## Identify and plot proteins with highest fold-change between conditions

Compute fold-change per gene between Sulforaphane and DMSO


```{r fold_change, echo=TRUE, fig.width=12.5, fig.height=12}
fold_changes <- protein_data_parsed %>%
    group_by(Gene, Label) %>%
    filter(n() >= 2) %>% # keep only labels with >=2 replicates
    ungroup() %>%
    group_by(Gene) %>%
    filter(n_distinct(Label) == 2) %>% # keep only proteins present in both groups
    summarize(
        DMSO = mean(Intensity[Label == "DMSO"], na.rm = TRUE),
        Sulforaphane = mean(Intensity[Label == "Sulforaphane"], na.rm = TRUE),
        .groups = "drop"
    ) %>%
    mutate(
        Log2FC = Sulforaphane - DMSO, # already log2-transformed
    ) %>%
    # Prepare a column indicating in which condition intensity is higher
    # This column will be used for faceting the plot later
    mutate(higher_in = case_when(
        Log2FC > 0 ~ "Higher in Sulforaphane",
        Log2FC < 0 ~ "Higher in DMSO",
        TRUE ~ "No change"
    )) %>%
    arrange(desc(Log2FC))
```

Extract top 20 proteins with highest absolute Log2FC in each condition

```{r}
top_proteins_DMSO <- fold_changes %>%
    filter(higher_in == "Higher in DMSO") %>%
    slice_max(order_by = abs(Log2FC), n = 20)
```


```{r}
top_proteins_sulforaphane <- fold_changes %>%
    filter(higher_in == "Higher in Sulforaphane") %>%
    slice_max(order_by = abs(Log2FC), n = 20)
```



Combine the top proteins from both conditions

```{r}
top_proteins_both <- bind_rows(top_proteins_DMSO, top_proteins_sulforaphane)
```


Prepare data for plotting

```{r}
plot_data <- protein_data_parsed %>%
    group_by(Gene, Label) %>%
    summarize(
        Mean = mean(Intensity, na.rm = TRUE),
        SD = sd(Intensity, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    # Inner join the current data frame with the top proteins to keep only those
    inner_join(top_proteins_both[, c("Gene", "higher_in")], by = "Gene")
```

Start ggplot, map global aesthetics (x and y axes, color, and group)
Also, reorder genes based on their mean intensity for better visualization
Add lines connecting the points of the same gene
(needs 'group' aesthetic for the correct behavior)

Add "shadow" points representing the standard deviation.
They should be behind the main points, so add them first.
Add points for each condition.
Refine size scale for the SD points for better visibility
Flip coordinates. What is mapped to x axis goes to y axis and vice versa.


```{r}
plot_data %>%
    ggplot(aes(x = reorder(Gene, Mean), y = Mean, color = Label, group = Gene)) +
    geom_line(color = "#999999", linewidth = 1, show.legend = FALSE) +

    geom_point(aes(size = SD),
        position = position_dodge(width = 0.5), alpha = 0.35,
        show.legend = TRUE
    ) +
    geom_point(position = position_dodge(width = 0.5), size = 4, show.legend = FALSE) +
    scale_size_continuous(range = c(6, 13), breaks = c(1, 2, 3)) +
    coord_flip() +
    labs(
        title = "Top 20 proteins with highest Log2FC between Sulforaphane and DMSO",
        x = "Gene symbol",
        y = "Mean Log2 [Intensity]",
        size = "Standard deviation",
        color = "Group"
    ) +
    scale_color_manual(values = label_colors) +
    theme_minimal() +
    facet_wrap(~higher_in, scales = "free_y", nrow = 1) +
    guides(color = guide_legend(
        override.aes = list(size = 5, alpha = 1)
    ))
```


