---
title: "Data Visualization in R"
subtitle: "Workshop material for 18th November 2025"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook
---
# Welcome
 
```{r,echo=FALSE,include=FALSE}
#  variables
rep_org <- "Data Science Platform" 
rep_core <- "Juliana Assis"
rep_core_email <- "jasge@dtu.dk"
rep_core2 <- "Andre Cunha"
rep_core_email2 <- "afecun@dtu.dk"
rep_core3 <- "Alberto Pallejà Caro"
rep_core_email3 <- "apca@biosustain.dtu.dk"

# Combine all responsibles into a single string
rep_all <- paste0(
  rep_core, " (", rep_core_email, ")<br>",
  rep_core2, " (", rep_core_email2, ")<br>",
  rep_core3, " (", rep_core_email3, ")"
)
```

<br>

::: boxy
__Organisation:__ `r rep_org`  
__Responsibles:__ `r paste0(rep_all," ")`  
:::

<br>
 


**Welcome to the Data Visualization with R Workshop!**


This workshop introduces **data visualization in R** using the simplest ggplot2, plotly.
We will cover both theory and practical exercises.

**Goals**

- Understand R visualization grammar
- Produce publication-ready plots
- Explore real datasets



**Slides**

``` {r libraries, echo=FALSE, eval=TRUE}
knitr::include_url("https://docs.google.com/presentation/d/1ifS-JNsKchnT3tlZtZsDes7oRqoKTbjO/embed")
```


Below are two setup options to the pratical activities:

>Run the workshop in the cloud (no need to install anything). 

**Launch the app:**

[![Open in GitHub Codespaces](https://github.com/codespaces/badge.svg)](https://codespaces.new/biosustain/dsp_workshop_datavizR)


>Run the workshop locally on your machine. 

To run it in your own machine install the following packages:

```{r, eval=FALSE}
# List of required CRAN packages
packages <- c(
  "dplyr",
  "ggplot2",
  "ggrepel",
  "RColorBrewer",
  "rmarkdown",
  "tidyr",
  "plotly"
)

# Install missing packages
install_if_missing <- function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg, repos = "https://cloud.r-project.org/")
  } else {
    message(pkg, " already installed.")
  }
}

# Run installation
invisible(lapply(packages, install_if_missing))

message("All requested R packages are installed and ready to use!")

```



**Local build of website**

```{r, eval=FALSE}
bookdown::render_book("index.Rmd", "bookdown::gitbook")
```




