---
title: "Data Visualization in R, `r format(Sys.Date(), '%d-%b-%Y')`"
subtitle: "JAG"
site: bookdown::bookdown_site
documentclass: book
output:
  bookdown::gitbook
---
 
 
# Welcome
 
Welcome to the Data Visualization with R Workshop! 


This workshop introduces **data visualization in R** using the simplest ggplot2, plotly.
We will cover both theory and practical exercises.

**Goals**

- Understand R visualization grammar
- Produce publication-ready plots
- Explore real datasets


Below are two setup options:

Run the workshop in the cloud (no need to install anything). 

**Launch the app:**

[![Open in GitHub Codespaces](https://github.com/codespaces/badge.svg)](https://github.com/biosustain/dsp_workshop_datavizR)



:::boxy
Run the workshop locally on your machine. 
:::


**Local build of website**

```{r, eval=FALSE}
bookdown::render_book("index.Rmd")
```



``` {r libraries, echo=FALSE, eval=TRUE}
knitr::include_url("https://docs.google.com/presentation/d/1ifS-JNsKchnT3tlZtZsDes7oRqoKTbjO/embed")
```



